# Pension Liability Calculations

## Present Value of Future Benefits (PVFB)
The actuarial present value of all expected future benefit payments:
```
PVFB = Σ expected_future_benefits * discount_factors * survival_probabilities
```

## Three Key Liability Measures

### 1. Market Liability
Market value using risk-adjusted discount rates matching traded securities with similar cash flow patterns.

### 2. Solvency Liability  
Present value using risk-free rates (e.g., Treasury bonds):
```
Solvency_Liability = Σ benefit_payments * treasury_discount_factors
```

### 3. Budget Liability (Actuarial Liability - AL)
Traditional actuarial accrued liability for funding purposes [@aaa_2004]:
```
AL = portion of PVFB attributed to past service
```

## Benefit Obligation Measures

### Projected Benefit Obligation (PBO)
Present value of benefits earned to date based on projected future salaries:
```
PBO_end = PBO_begin + Service_Cost + Interest_Cost ± Actuarial_Gains_Losses + Plan_Amendments - Benefits_Paid
```

```{r}
#| label: pbo-calculation
# PBO calculation function
calculate_pbo_end <- function(
  pbo_begin,
  service_cost,
  interest_cost,
  actuarial_gains_losses,
  plan_amendments,
  benefits_paid
) {
  pbo_end <- pbo_begin +
    service_cost +
    interest_cost +
    actuarial_gains_losses +
    plan_amendments -
    benefits_paid
  return(pbo_end)
}
```

### Service Cost
Present value of benefits earned in current year:
```
Service_Cost = present_value_of_current_year_benefit_accrual
```

### Interest Cost
Growth in obligation due to passage of time:
```
Interest_Cost = PBO_beginning * discount_rate
```

```{r}
#| label: interest-cost
# Interest cost calculation
calculate_interest_cost <- function(pbo_beginning, discount_rate) {
  return(pbo_beginning * discount_rate)
}
```